---
title: "Data Cleaning for Occupational Illness"
author: "Saritha Kumari Krishna Reddy, Stephanie Ajah, Karanvir Virdi"
date: "Dec 12, 2024"
format: html
execute: 
  embed-resources: true
---

## Introduction

The file focuses on processing occupational illness data spanning multiple years (2016–2022). The `tidyverse` and `readxl` packages are employed for data manipulation and importing excel files. The `read_xlsx` function loads datasets for each year, which are subsequently combined into a single dataset for further analysis.

Key cleaning steps include mapping NAICS codes to industries, converting state abbreviations to full names in lower case, and addressing missing company names. The script selects relevant variables, renames columns for clarity, and prepares the data for analysis. Finally, the cleaned data is saved as a CSV file for downstream tasks.

## Step 1: Load required libraries and files

The library 'tidyverse' is used in this assignment because it holds all the core packages used for data manipulation, data cleaning and visualization like dplyr, stringr, ggplot2 and many more.

```{r message=FALSE}
# Run the following commands to install the required packages, if you do not have the packages installed
# install.packages("tidyverse")
# install.packages("readxl")

library(tidyverse)
library(readxl)
```

To read xlsx file we used the function `read_xlsx` from `readxl` package by passing the file path as the required parameter.

```{r warning=FALSE}
occ_data_2016 <- read_xlsx('/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset_Occ/Data_2016.xlsx')
occ_data_2017 <- read_xlsx('/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset_Occ/Data_2017.xlsx')
occ_data_2018 <- read_xlsx('/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset_Occ/Data_2018.xlsx')
occ_data_2019 <- read_xlsx('/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset_Occ/Data_2019.xlsx')
occ_data_2020 <- read_xlsx('/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset_Occ/Data_2020.xlsx')
occ_data_2021 <- read_xlsx('/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset_Occ/Data_2021.xlsx')
occ_data_2022 <- read_xlsx('/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset_Occ/Data_2022.xlsx')
```

## Step 2: Data Cleaning and Data Transformation

-   Identifying and fixing errors, inconsistencies, or missing values. Ensuring that the data is accurate, complete, and suitable for processing.

-   Reshaping and modifying the cleaned data to make it suitable for analysis, visualization, or modeling. Like creating new columns, aggregating data, or converting data into meaningful formats.

The following piece of code combines datasets from 2016 to 2022 into a single data frame.

```{r warning=FALSE}
occ_data <- rbind(occ_data_2016, occ_data_2017, occ_data_2018, occ_data_2019, occ_data_2020, occ_data_2021, occ_data_2022)
```

The code snippet categorizes industries based on `Primary NAICS` codes into broader sectors using `case_when()` and regex pattern matching. It maps each NAICS code to a specific level industry category and assigns "Other" for codes that don't match predefined patterns.

```{r warning=FALSE}
# PRIMARY NAICS
occ_data <- occ_data |>
  rowwise() |>
  mutate(industry = case_when(
    str_detect(`naics_code`, regex("^11", ignore_case = TRUE)) ~ "Agriculture, Forestry, Fishing and Hunting",
    str_detect(`naics_code`, regex("^21", ignore_case = TRUE)) ~ "Mining, Quarrying, and Oil and Gas Extraction",
    str_detect(`naics_code`, regex("^22", ignore_case = TRUE)) ~ "Utilities",
    str_detect(`naics_code`, regex("^23", ignore_case = TRUE)) ~ "Construction",
    str_detect(`naics_code`, regex("^31|32|33", ignore_case = TRUE)) ~ "Manufacturing",
    str_detect(`naics_code`, regex("^42", ignore_case = TRUE)) ~ "Wholesale Trade",
    str_detect(`naics_code`, regex("^44|45", ignore_case = TRUE)) ~ "Retail Trade",
    str_detect(`naics_code`, regex("^48|49", ignore_case = TRUE)) ~ "Transportation and Warehousing",
    str_detect(`naics_code`, regex("^51", ignore_case = TRUE)) ~ "Information",
    str_detect(`naics_code`, regex("^52", ignore_case = TRUE)) ~ "Finance and Insurance",
    str_detect(`naics_code`, regex("^53", ignore_case = TRUE)) ~ "Real Estate and Rental and Leasing",
    str_detect(`naics_code`, regex("^54", ignore_case = TRUE)) ~ "Professional, Scientific, and Technical Services",
    str_detect(`naics_code`, regex("^56", ignore_case = TRUE)) ~ "Administrative and Support and Waste Management and Remediation Services",
    str_detect(`naics_code`, regex("^61", ignore_case = TRUE)) ~ "Educational Services",
    str_detect(`naics_code`, regex("^62", ignore_case = TRUE)) ~ "Health Care and Social Assistance",
    str_detect(`naics_code`, regex("^71", ignore_case = TRUE)) ~ "Arts, Entertainment, and Recreation",
    str_detect(`naics_code`, regex("^72", ignore_case = TRUE)) ~ "Accommodation and Food Services",
    str_detect(`naics_code`, regex("^81", ignore_case = TRUE)) ~ "Other Services (except Public Administration)",
    str_detect(`naics_code`, regex("^92", ignore_case = TRUE)) ~ "Public Administration",
    TRUE ~ "Other"
  ))
```

The below code snippet maps state abbreviations to their full names and standardize them to lowercase.

```{r warning=FALSE}
# Convert all states to lowercase
state_abbreviation_to_name <- c(
  "al" = "alabama", "ak" = "alaska", "az" = "arizona", "ar" = "arkansas",
  "ca" = "california", "co" = "colorado", "ct" = "connecticut", "de" = "delaware",
  "fl" = "florida", "ga" = "georgia", "hi" = "hawaii", "id" = "idaho",
  "il" = "illinois", "in" = "indiana", "ia" = "iowa", "ks" = "kansas",
  "ky" = "kentucky", "la" = "louisiana", "me" = "maine", "md" = "maryland",
  "ma" = "massachusetts", "mi" = "michigan", "mn" = "minnesota", "ms" = "mississippi",
  "mo" = "missouri", "mt" = "montana", "ne" = "nebraska", "nv" = "nevada",
  "nh" = "new hampshire", "nj" = "new jersey", "nm" = "new mexico", "ny" = "new york",
  "nc" = "north carolina", "nd" = "north dakota", "oh" = "ohio", "ok" = "oklahoma",
  "or" = "oregon", "pa" = "pennsylvania", "ri" = "rhode island", "sc" = "south carolina",
  "sd" = "south dakota", "tn" = "tennessee", "tx" = "texas", "ut" = "utah",
  "vt" = "vermont", "va" = "virginia", "wa" = "washington", "wv" = "west virginia",
  "wi" = "wisconsin", "wy" = "wyoming",
  "pr" = "puerto rico", "vi" = "virgin islands", "gu" = "guam", "pw" = "unknown",
  "mp" = "unknown", "as" = "unknown", "aa" = "unknown", "ae" = "unknown", "ap" = "unknown"
)

# Convert state abbreviations in the dataset to lowercase
occ_data$state <- tolower(occ_data$state)

# Map state abbreviations to full names
occ_data$state <- state_abbreviation_to_name[occ_data$state]

# Remove rows where state is "unknown"
occ_data <- occ_data[occ_data$state != "unknown", ]
```

The following code replaces missing company names with establishment names to deal with the missing values in the column `company_name`.

```{r}
occ_data$company_name <- ifelse(is.na(occ_data$company_name), occ_data$establishment_name, occ_data$company_name)
```

The code cleans and selects relevant columns from the occ_data dataset, making it more manageable for analysis. By renaming the columns, it improves both the clarity and the quality of the data for further processing.

```{r}
occ_data <- occ_data |>
  select(
    year_filing_for,
    company_name,
    state,
    industry, 
    annual_average_employees, 
    total_hours_worked,
    no_injuries_illnesses, 
    total_deaths,
    total_dafw_cases,
    total_djtr_cases,
    total_other_cases,
    total_dafw_days, 
    total_djtr_days,
    total_injuries, 
    total_poisonings,
    total_respiratory_conditions,
    total_skin_disorders,
    total_hearing_loss,
    total_other_illnesses
  ) |>
  rename(
    year = year_filing_for,
    injury_illness = no_injuries_illnesses
  )
```

## Step 3: Saving the clean data into a file

The code saves the processed occ_data dataframe to a CSV file at the specified file_path.

```{r}
file_path <- "/Users/sarithakumarik/Documents/DATA6200/FinalProject/Dataset/occupational_illness_data.csv"
write.csv(occ_data, file_path, row.names = FALSE)
```